---
title: 'Lab 9: Fixed vs Random Effects'
embed-resources: true
author: "Xiuyu Cao"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
format:
  html:
    code-folding: show
    highlight: textmate
    number-sections: true
    theme: flatly
    toc: TRUE
    toc-depth: 4
    toc-float:
      collapsed: false
      smooth-scroll: true
---

## Lab Setup
Set your working directory and load libraries.
```{r}
library(nlme)  # for running mixed models
```

# Exercise 1
## Question 1.1
**Answer:** In a previous research project I have worked on, I was thinking about modeling microclimate (fine-resolution temperature) using macroclimate (coarser-resolution temperature) and land cover type (`micro_temp ~ macro_temp + land_cover`). However, there may be potential omitted variables introducing bias to my model. For example, the distance to the center of urban area may influence both the independent variable `macro_temp` and dependent variable `micro_temp`, since the Urban Heat Island effect would cause the averaged large scale temperature `macro_temp` to raise, as well as the small scale `micro_temp`.


# Exercise 2
```{r}
irrdata <- read.csv('../data/lab9/irrdata.csv')
head(irrdata, 3)
```

* `irrigation`: the number of times a farmer irrigated his/her crop throughout the growing season
* `yield`: the yield of the wheat crop (in kg/ha)
* `state`: the State the farmer lives in

## Question 2.1
```{r}
lmod3 = lm(yield ~ irrigation, data = irrdata)
summary(lmod3)

# state farmer live in --> the wealth of the farmer
table(irrdata$state)

lmod4 = lm(yield ~ irrigation + state, data = irrdata)
summary(lmod4)
```
**Answer:**

* *Differences:* 
  * The beta coefficients are not the the same
  * Second model's F-statistic decreases
* *Similarities:*
  * What is significant: In both models, the intercept and slope for `irrigation` are significant
  * $R^2$ being similar
  * Residual standard error being similar


## Question 2.2
**Answer:** The beta coefficient on `irrigation` decreased a bit when including `state` in the model because `state` explains some degrees of the yield, due to the difference in wealth of farmers from different states. And since the farmers in Punjab are much richer than those from Bihar, `statePunjab` has a positive intercept difference, though not significant.


# Exercise 3
```{r}
lakedf <- read.csv('../data/lab9/lakedf.csv')
head(lakedf, 3)
```
* `lake`: the number ID of the lake where the sample was taken
* `diversity`: the number of different algae species
* `prod`: the productivity of the algae (in kg/m)
* `temp`: mean temperature for 2017 (in Celsius)
* `rainfall`: total rainfall in 2017 (in mm)
* `sun`: the amount of sunlight measured as insolation (in W/m2)

## Question 3.1
```{r}
lmod5 = lm(prod ~ diversity + temp + rain + sun, data = lakedf)
summary(lmod5)

# add random effect
mod = lme(prod ~ diversity + temp + rain + sun, random = ~1|lake, data = lakedf)
summary(mod)
```
**Answer:** The beta coefficient of the continuous variables are similar, as well as the p values of the continuous variables. While for the intercept, although the p values of the two models are similar, the beta coefficient of the intercept is different, due to the second model's intercept being added a random effect.

## Question 3.2
**Answer:** In a previous course project, I was trying to model income of a place using variables including the distance to parks, forests, and water, density of roads, elevation, and population. I was trying to get a generalized model that can get income across the US, instead of only the few states whose data I have access to. Also, in this case I believe I have collected enough factors and I don't think there is a problem of OVB. Therefore, in this case a random effect on the states would be more appropriate than a fixed effect.

