---
title: 'Lab 8: ANCOVA and Interactions'
embed-resources: true
author: "Xiuyu Cao"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
format:
  html:
    code-folding: show
    highlight: textmate
    number-sections: true
    theme: flatly
    toc: TRUE
    toc-depth: 4
    toc-float:
      collapsed: false
      smooth-scroll: true
---

## Lab Setup
```{r, message=F}
# Load libraries
library(dplyr)  # for data manipulation
library(car)  # for equal variance test
library(lmtest)  # for lm assumption test

# Read in data
dataset <- read.csv("../data/lab8/sanddata.csv", sep = ",", header = T,comment.char = "#", stringsAsFactors = T)
head(dataset, 3)
```
* `juveniles`: number of juvenile cranes counted annually
* `sand`: tons of sand removed annually
* `temperature`: mean annual temperature in degrees Celsius
* `rainfall`: total annual rainfall in mm
* `humanpop`: total human population (in 1000s) within 50 miles of each site
* `country`: country where each site is found (China, India, Indonesia, or Malaysia)


# Exercise 1
```{r}
# check multicollinearity
cor(dataset[, -6], dataset[,-6])  # get corr, exclude col 6

# get linear model
linmod <- lm(juveniles ~ sand + temperature + rainfall + humanpop + country, data=dataset)

# check linear relationship
par(mfrow = c(2,2))
plot(juveniles ~ sand, data = dataset)
plot(juveniles ~ temperature, data = dataset)
plot(juveniles ~ rainfall, data = dataset)
plot(juveniles ~ humanpop, data = dataset)

# check residual homoscedasticity and normality
par(mfrow = c(2,2))
plot(linmod)

# get results
summary(linmod)
```

## Question 1.1
**Answer:**

*Interpretation of the significant coefficients (at $\alpha=0.05$)*

* Intercept ($P=2\times10^{-16}<\alpha$): With `sand`, `temperature`, `rainfall`, and `humanpop` being zero, it is predicted that there will be 4603.66 juvenile cranes on average in China.
* sand ($P=2\times10^{-16}<\alpha$): Every one ton of sand removed is associated with a reduction of 1.5 juvenile cranes (copied from lab instruction).
* temperature ($P=1.73\times10^{-8}<\alpha$): Every degree Celsius of temperature raised is associated with a reduction of 76.8 juvenile cranes.
* humanpop ($P=0.005<\alpha$): Every 1,000 increase in population within 50 miles is associated with a reduction of 0.4 juvenile cranes.
* countryMalaysia ($P=0.0004<\alpha$): On average, Malaysia has 190.5 fewer juvenile cranes than China when controlling for `sand`, `temperature`, `rainfall`, and `humanpop`.

## Question 1.2
**Answer**: 

*Interpretation of the non-significant coefficients (at $\alpha=0.05$)*

* rainfall ($P=0.43>\alpha$): There is no significant linear relationship between `rainfall` and `juveniles`.
* countryIndia ($P=0.53>\alpha$): There is no significant difference in `juveniles` between India and China, controlling for `sand`, `temperature`, `rainfall`, and `humanpop`.
* countryIndonesia ($P=0.38>\alpha$): There is no significant difference in `juveniles` between India and China, controlling for `sand`, `temperature`, `rainfall`, and `humanpop`.


# Exercise 2
## Question 2
```{r}
table(dataset$country)

# use India as baseline intercept
dataset.releveled <- dataset %>%
  mutate(country=relevel(country, ref='India'))  # relevel India to the first

# rerun the linear model
linmod2 <- lm(juveniles ~ sand + temperature + rainfall + humanpop + country, data = dataset.releveled)
summary(linmod2)
```
**Answer**:

Coefficients of the continuous variables (`sand`, `temperature`, `rainfall`, and `humanpop`) stay the same while those of the categorical ones and the intercept are changed. This is due to the change of the baseline variable (China to India). Now the intercept is the cranes in India, instead of China, when the continuous variables are zero, and the coefficients of the categorical variables are now the difference of them from the `juveniles` in India.



# Exercise 3 
```{r warnings=F}
# get model
anomod <- aov(juveniles ~ sand + temperature + rainfall + humanpop + country, data = dataset)

# check the assumption of equal variance
leveneTest(juveniles ~ country, data = dataset)  # from the result, met

# get model results
summary(anomod)

# do Tukey HSD to see which groups are different
tukey.mat <- TukeyHSD(anomod)$country
tukey.mat
```
## Question 3
**Answer**:

*Interpretation of the Tukey HSD test result:*

We can get additional information on pairwise comparisons instead of only the countries comparing to the baseline country when doing `lm()`. The countries that are significantly different from one another are as follows.
```{r get significantly different pairs from Tukey HSD result}
i.significant <- tukey.mat[,4]<0.05
tukey.mat[i.significant,]
```

# Exercise 4
## Question 4
```{r}
# change the order of the variables
anomod3 <- aov(juveniles ~ temperature + rainfall + humanpop + country + sand, data = dataset)
summary(anomod3)

```
**Answer**: 

Comparing the p values between `anomod` and `anomod3`. Which variables are significant (at $\alpha=0.05$) are the same (only `rianfall` not being significant). However, the p values are not the same between these two models.


# Exercise 5
```{r}
linmod5 <- lm(juveniles ~ sand + temperature + rainfall + humanpop + country + humanpop*country, data = dataset)
summary(linmod5)

# visualize
plot(1:1000, rep(-100, 1000), ylim = c(4000, 5500), xlim = c(50, 1000), xlab = 'Human Population', ylab = 'Juveniles')
abline(a = 5054.11977, b = -1.24563, col = 'blue') # China
abline(a = 5054.11977 - 818.65030, b = -1.24563 + 1.31526, col = 'red') # India
abline(a = 5054.11977 - 961.39246 , b = -1.24563 + 1.53082, col = 'orange') # Indonesia
abline(a = 5054.11977 - 676.65413, b = -1.24563 + 0.81239, col = 'green') # Malaysia
legend("topright", legend=c("China", "India", "Indonesia", "Malaysia"),
       col=c("blue", "red", "orange", "green"), lty=1)
```
## Question 5
**Answer**: 

The numbers to calculate the slopes for each country are from the `humanpop` and `humanpop:countryIndia`, `humanpop:countryIndonesia`, and `humanpop:countryMalaysia`. The latter three are the difference in slopes from the baseline slope (slope of `humanpop`).


# Extra Credit (Code and outputs required)
## Check Data
```{r}
iris <- iris %>%
  mutate(Species=as.factor(Species))
head(iris, 3)
```
I would like to study the relationship of `Petal.Width` between other variables.

## Check Multicollinearity
```{r}
# plot relationship
pairs(iris[,-5])  # scatter plot of the continuous variables

# check for multicollinearity
cor(iris[,-5], iris[,-5])
```
The correlation coefficient between `Petal.Length` and `Sepal.Length` (0.87) is high, indicating a strong positive relationship, which may cause multicollinearity. Therefore I would exclude `Petal.Length`.

## Check Linear Regression Assumptions
```{r}
mod <- lm(Petal.Width~Sepal.Length+Sepal.Width+Species, data=iris)
summary(mod)

# linear relationship
par(mfrow = c(1,2))
plot(Petal.Width ~ Sepal.Length, data=iris)
plot(Petal.Width ~ Sepal.Width, data=iris)

# Homoscedasticity
bptest(mod)

# Error Normality
shapiro.test(residuals(mod))
```
*Linear Regression Assumptions:*

* Linear relationship between variables: from the plot - met.
* Statistical independence of the errors: assume so.
* Homoscedasticity: $P=0.0003<\alpha=0.05$ --> reject null hypothesis of homoscedasticity - not met.
* Error Normality: $P=0.29>\alpha=0.05$  --> fail to reject null hypothesis of error normality - met.

## Results
```{r}
summary(mod)
```
*Result Interpretation:*

* Intercept ($P=9.73\times10^{-7}<\alpha=0.05$): with `Sepal.Length` and `Sepal.Width` being zero, it is predicted that `Petal.Width` of species *Setosa* is -0.86897 (not meaningful in real life context though).
* `Sepal.Length` ($P=0.06>\alpha=0.05$): no significant relationship between `Sepal.Length` and `Petal.Width`.
* `Sepal.Width` ($P=1.29\times10^{-5}<\alpha=0.05$): every one unit increase in `Sepal.Width` is associated with 0.23 unit increase in `Petal.Width`.
* `Speciesversicolor` ($P=2\times10^{-16}<\alpha=0.05$): on average, species *Versicolor* has 1.17 unit more in `Petal.Width` than species *Setosa*, controlling for `Sepal.Length` and `Sepal.Width`.
* `Speciesvirginica` ($P=2\times10^{-16}<\alpha=0.05$): on average, species *Virginica* has 1.78 unit more in `Petal.Width` than species *Setosa*, controlling for `Sepal.Length` and `Sepal.Width`.

