---
title: "Lab 6: Linear Regression"
embed-resources: true
author: "Xiuyu Cao"
date: "`r format(Sys.Date(), '%m/%d/%Y')`"
format:
  html:
    code-folding: show
    highlight: textmate
    number-sections: true
    theme: flatly
    toc: TRUE
    toc-depth: 4
    toc-float:
      collapsed: false
      smooth-scroll: true
---

# Exercise 1
```{r}
pairs(airquality[,c('Ozone','Solar.R', 'Wind', 'Temp')])  # scatter plot

cor(airquality[, c("Ozone", "Solar.R", "Wind", "Temp")], 
    use = "complete.obs")    # use the rows where there is both a value for x and y
```
## Question 1.1
**Answer:** The correlation between `Ozone` and `Wind` is -0.6124966. In my opinion, this is a moderate to strong *negative* correlation. It means when ozone level increases, wind speed decreases, and vice versa.

## Question 1.2
**Answer:** The correlation between `Ozone` and `Solar.R` is 0.3483417. In my opinion, this is a weak to moderate *positive* correlation. It means when the ozone level increases, Solar R increases, and vice versa.


# Exercise 2
## Question 2.1
```{r}
plot(eruptions~waiting, data=faithful)  # Plot waiting against eruptions
```

## Question 2.2
```{r}
# Calculate the correlation coefficient
cor(faithful$eruptions, faithful$waiting)
```


# Exercise 3
## Question 3.1
Notes: For full points, include model, interpretation and plot.

For a complete interpretation, include slope significance (alpha and p-value), slope and R2 interpretation, and intercept value with alpha and p-value.
```{r}
# Build linear model
mod <- lm(Temp~Ozone, data=airquality)
summary(mod)

# Plot scatterplot with estimated regression line
plot(Temp~Ozone, data=airquality)
abline(mod, col='red')
```

**Answer:**

* The slope is 0.20081. This means for every 1 unit increase in ozone, the temperature increases by 0.20081. The P value of the slope is $2 \times 10^{-16}<<\alpha=0.05$. So we reject $H_0:Slope=0$, there is a significant relationship between ozone and temperature.
* $R^2=0.4877$. 48.77% of the variation in temperature is explained by ozone.
* The intercept is 69.41072. This means that when ozone is zero, temperature is predicted to be 69.41072. The intercept has a p value of $2 \times 10^{-16}<<\alpha=0.05$. Meaning the temperature is significantly different from zero when ozone is zero.

## Question 3.2
Note: For full points, include model, interpretation and plot.
```{r}
# Build linear model
mod <- lm(eruptions~waiting, data=faithful)
summary(mod)

#Plot scatterplot with estimated regression line
plot(eruptions~waiting, data=faithful)
abline(mod, col='red')
```

**Answer:**

* The slope is 0.075628. This means for every 1 unit increase in waiting time, the eruption time increases by 0.075628. The P value of the slope is $2 \times 10^{-16}<<\alpha=0.05$. So we reject $H_0:Slope=0$, there is a significant relationship between waiting and eruptions.
* $R^2=0.8115$. 81.15% of the variation in eruptions is explained by waiting time.
* The intercept is -1.874016. This means that when waiting time is zero, eruption time is predicted to be -1.87 (not meaningful in real world context). The intercept has a p value of $2 \times 10^{-16}<<\alpha=0.05$. Meaning the eruption time is significantly different from zero when waiting time is zero.


# Exercise 4

## Question 4.1
Note: To receive full credit, answer the three parts.
```{r}
mod <- lm(Temp~Wind, data=airquality)
r2 <- summary(mod)$r.squared

# get SSY
y <- airquality$Temp
meany <- mean(y)
diff <- y-meany
SSY <- sum(diff^2)

# get SSYp
x <- data.frame(Wind=airquality$Wind)
predy <- predict.lm(mod, x)
diff <- predy-meany
SSYp <- sum(diff^2)

# get SSE
diff <- predy-y
SSE <- sum(diff^2)
SSE

# Verify SSY - SSYp = SSE
SSY-SSYp==SSE
```

**What is SSE measuring in non-technical terms?**

SSE is the sum of the squared deviations between the actual value and the predicted value. It's measuring the amount of vairation not explained by the model.


# Exercise 5
## Question 5.1
Note: Your final answer should include three parts: the model, tests for assumptions of homoscedasticity (perform all 3) and normality (perform both shapiro.test and qqPlot) and interpret of the results of homoscedasticity and normality tests.
```{r, message=F}
library(lmtest)
library(car)
library(dplyr)

# remove NAs
dat.airquality <- airquality %>%
  select(Temp, Ozone)  %>%
  na.omit()

# Create linear model
mod <- lm(Temp~Ozone, data=dat.airquality)
summary(mod)
```

### Check whether the model meets assumptions
#### Residual Homoscedasticity
```{r}
# visual
plot(residuals(mod)~Ozone, data=dat.airquality,
     main='Residuals v.s. Fitted', xlab='Fitted', ylab='Residuals')
abline(h=0, col='red')

# statistical test
bptest(mod)
```
**Result Interpretation:** The p value of the Breush-Pegan test is $0.0721>\alpha=0.05$. We thus fail to reject $H_0$: The variance is constant. Therefore, the requirement of homoscedasticity is met.

#### Normal Distributed Residuals
```{r}
# Q-Q plot
qqPlot(residuals(mod))

# Shapiro-Wilk test
shapiro.test(residuals(mod))
```
**Result Interpretation:** From the Q-Q plot, some points laid out of the intervals. From the Shapiro-Wilk test, the p value is $0.000226<\alpha=0.05$. We thus reject $H_0$: the data being tested are drawn from a normally distributed population. Therefore, the requirement of residual normality is NOT met.


